Towards a Mechanized Proof of Noninterference

### 2013-12-18 paper/paper.tex
Two columns is more comfortable to my eye for some reason...
More futzing with formatting.
Made todo's easier to pick out of paper.
Fixed some lemma labels.

\documentclass[a4paper]{article}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
%\usepackage{float}
%\floatstyle{boxed}
%\restylefloat{figure}

\newcommand{\langName}[0]{TinyML\textsuperscript{2}}
\newcommand{\origLang}[0]{CoreML\textsuperscript{2}}
\newcommand{\typeRule}[3]{#1 \vdash #2 \colon #3}
\newcommand{\guards}[0]{\lhd}
\newcommand{\lift}[1]{\lfloor #1 \rfloor}
\newcommand{\todo}[1]{\textbf{TODO: #1}}

@@

\begin{figure}[h!]
  \begin{align}
    v & ::= x \: | \: k \: | \: \textsf{fix} f . \lambda x . e \tag*{ } \\
    v' & ::= v \: | \: <v | v> \tag{values}\\
    a & ::= v \tag*{ } \\
    a' & ::= a \: | \: <a | a> \tag{answers} \\
    e & ::= a \: | \: v v \: | \: \textsf{if } v \: e \: e \: | \: \textsf{let
    } x := v \textsf{ in } e \tag*{ } \\
    e' & ::= e \: | \: <e | e> \tag{expressions}
  \end{align}
  \caption{Syntax of \langName.  The prime variants are the full class of
  expressions, values, and answers.  This construction avoids allowing nested
  branching values.}
  \label{fig:syntax}
\end{figure}

@@

\begin{figure}[h!]
  Basic reductions
  \begin{align}
    (\textsf{fix} f . \lambda x . e) v
    \: \to \: &
    e[x \Leftarrow v][f \Leftarrow \textsf{fix} f . \lambda x . e]
    \tag{$\beta$}
    \label{eq:beta}
    \\
    \textsf{if } 1 \: e_1 \textsf { } e_2
    \: \to \: &
    e_1
    \tag{if-true}
    \label{eq:if-true}
    \\
    \textsf{if } (k \neq 1) \textsf { } e_1 \textsf { } e_2
    \: \to \: &
    e_2
    \tag{if-false}
    \label{eq:if-false}
    \\
    \textsf{let }  x := v \textsf{ in } e
    \: \to \: &
    e[x \Leftarrow v]
    \tag{let}
    \label{eq:let}
  \end{align}

  Lifting
  \begin{align}
    < v_1 | v_2 > v
    \: \to \: &
    < v_1 \lift{v}_1 | v_2 \lift{v}_2 >
    \tag{lift-app}
    \label{eq:lift-app}
    \\
    \textsf{if } < v_1 | v_2 > \: e_1 \: e_2
    \: \to \: &
    < \textsf{if } v_1 \: \lift{e_1}_1 \: \lift{e_2}_1 \: |
    \textsf{ if } v_2 \: \lift{e_1}_2 \: \lift{e_2}_2 >
    \tag{lift-if}
    \label{eq:lift-if}
    \\
    &
    \infer*
    {
    e_1 \to e_1' \\
    }
    {
    <e_1 | e_2> \: \to \: <e_1' | e_2>
    }
    \tag{bracket-left}
    \label{eq:bracket-left}
    \\
    &
    \infer*
    {
    e_2 \to e_2' \\
    }
    {
    <e_1 | e_2> \: \to \: <e_1 | e_2'>
    }
    \tag{bracket-right}
    \label{eq:bracket-right}
  \end{align}
  \caption{Operational semantics for \langName.}
  \label{fig:operational_semantics}
\end{figure}

@@

\begin{figure}[h!]
  \begin{align*}
    t ::= int^l \: | \: (t -> t)^l
  \end{align*}
  \caption{Types for \langName}
  \label{fig:types}
\end{figure}

@@

\begin{figure}[h!]
  \begin{mathpar}
    \infer[S-Int]
    {
    l \leq l'
    }
    {
    \textsf{int}^l \leq \textsf{int}^{l'}
    }

@@

\begin{figure}[h!]
  \begin{mathpar}
    \infer[G-Int]
    {
    l \leq l'
    }
    {
    l \guards \textsf{int}^{l'}
    }

@@

\begin{figure}[h!]
  \begin{mathpar}
    \infer[V-Int]
    { }
    {\typeRule{\Gamma}{k}{\textsf{int}^{*}}}

@@

    \infer[V-Sub]
    {
    \typeRule{\Gamma}{v}{t'} \\
    t' \leq t
    }
    {\typeRule{\Gamma}{v}{t}}
  \end{mathpar}
  \caption{Typing judgements for values in \langName.}
  \label{fig:value_types}
\end{figure}

@@

\begin{figure}[h!]
  \begin{mathpar}
    \infer[E-Value]
    {
    \typeRule{\Gamma}{v}{t}
    }
    {
    \typeRule{*, \Gamma}{v}{t}
    }

@@

    \infer[E-Bracket]
    {
    \typeRule{pc \sqcup pc', \Gamma}{e_1}{t} \\
    \typeRule{pc \sqcup pc', \Gamma}{e_2}{t} \\
    pc' \in H \\
    pc' \guards t
    }
    {
    \typeRule{pc, \Gamma}{<e_1 | e_2>}{t}
    }
  \end{mathpar}
  \caption{Typing judgements for expressions in \langName.}
  \label{fig:expr_types}
\end{figure}

@@

  As a result, we know that $e$ reduces to an answer, because otherwise it would
  be stuck and this would mean that one of the projections is also stuck by
  Lemma~\ref{lem:bracket_stuck}, a contradicton.  Thus, we know that if both
  projections step to an answer, so must the original expression.
\end{proof}

@@

\begin{lemma}[Substitution]
  \label{lem:stuck_bracket}
  Assume $\typeRule{ }{v}{s}$.  Then, $\typeRule{\Gamma[x \mapsto s]}{v'}{t}$
  implies $\typeRule{\Gamma}{v'[x \Leftarrow v]}{t}$.  Also $\typeRule{pc,
  \Gamma[x \mapsto s]}{e}{t}$ implies $\typeRule{pc, \Gamma}{e[x \Leftarrow
  v]}{t}$.
\end{lemma}
\begin{proof}
  We can prove both statements by induction on the typing derivation for $v'$
  and $e$.  In this proof, we focus only on a subset of the cases.

@@

Our proof of Lemma~\ref{lem:soundness} actually turned out to be instructive
because it forced us to recognize that the original formulation in Simonet and
Pottier's paper on \origLang~\cite{InfoFlowML} wasn't quite correct.
Specifically, it is entirely possible that a step in the operational semantics
does not translate to a step in the operational semantics under either
projection, as in the case of the \ref{eq:lift-app} rule, where $\lift{e}_i$ and
$\lift{e'}_i$ are identical.  Luckily, we are fairly confident, given that we
were still able to prove Lemma~\ref{lem:bracket_stuck} despite this change, that
this is not a critical error in the original paper.

>>>>

\documentclass[a4paper,twocolumn]{article}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{color}
%\usepackage{float}
%\floatstyle{boxed}
%\restylefloat{figure}

\newcommand{\langName}[0]{TinyML\textsuperscript{2}}
\newcommand{\origLang}[0]{CoreML\textsuperscript{2}}
\newcommand{\typeRule}[3]{#1 \vdash #2 \colon #3}
\newcommand{\guards}[0]{\lhd}
\newcommand{\lift}[1]{\lfloor #1 \rfloor}
\newcommand{\todo}[1]{\textbf{\textcolor{red}{TODO: #1}}}

@@

\begin{figure}[t]
  \begin{align}
    v & ::= x \: | \: k \: | \: \textsf{fix} f . \lambda x . e \tag*{ } \\
    v' & ::= v \: | \: <v | v> \tag{values}\\
    a & ::= v \tag*{ } \\
    a' & ::= a \: | \: <a | a> \tag{answers} \\
    e & ::= a \: | \: v v \: | \: \textsf{if } v \: e \: e \: | \: \textsf{let
    } x := v \textsf{ in } e \tag*{ } \\
    e' & ::= e \: | \: <e | e> \tag{expressions}
  \end{align}
  \caption{Syntax of \langName.  The prime variants are the full class of
  expressions, values, and answers.  This construction avoids allowing nested
  branching values.}
  \label{fig:syntax}
\end{figure}

@@

\begin{figure*}[t]
  Basic reductions
  \begin{align}
    (\textsf{fix} f . \lambda x . e) v
    \: \to \: &
    e[x \Leftarrow v][f \Leftarrow \textsf{fix} f . \lambda x . e]
    \tag{$\beta$}
    \label{eq:beta}
    \\
    \textsf{if } 1 \: e_1 \textsf { } e_2
    \: \to \: &
    e_1
    \tag{if-true}
    \label{eq:if-true}
    \\
    \textsf{if } (k \neq 1) \textsf { } e_1 \textsf { } e_2
    \: \to \: &
    e_2
    \tag{if-false}
    \label{eq:if-false}
    \\
    \textsf{let }  x := v \textsf{ in } e
    \: \to \: &
    e[x \Leftarrow v]
    \tag{let}
    \label{eq:let}
  \end{align}

  Lifting
  \begin{align}
    < v_1 | v_2 > v
    \: \to \: &
    < v_1 \lift{v}_1 | v_2 \lift{v}_2 >
    \tag{lift-app}
    \label{eq:lift-app}
    \\
    \textsf{if } < v_1 | v_2 > \: e_1 \: e_2
    \: \to \: &
    < \textsf{if } v_1 \: \lift{e_1}_1 \: \lift{e_2}_1 \: |
    \textsf{ if } v_2 \: \lift{e_1}_2 \: \lift{e_2}_2 >
    \tag{lift-if}
    \label{eq:lift-if}
    \\
    &
    \infer*
    {
    e_1 \to e_1' \\
    }
    {
    <e_1 | e_2> \: \to \: <e_1' | e_2>
    }
    \tag{bracket-left}
    \label{eq:bracket-left}
    \\
    &
    \infer*
    {
    e_2 \to e_2' \\
    }
    {
    <e_1 | e_2> \: \to \: <e_1 | e_2'>
    }
    \tag{bracket-right}
    \label{eq:bracket-right}
  \end{align}
  \caption{Operational semantics for \langName.}
  \label{fig:operational_semantics}
\end{figure*}

@@

\begin{figure}[t]
  \begin{align*}
    t ::= int^l \: | \: (t -> t)^l
  \end{align*}
  \caption{Types for \langName}
  \label{fig:types}
\end{figure}

@@

\begin{figure}[t]
  \begin{mathpar}
    \infer[S-Int]
    {
    l \leq l'
    }
    {
    \textsf{int}^l \leq \textsf{int}^{l'}
    }

@@

\begin{figure}[t]
  \begin{mathpar}
    \infer[G-Int]
    {
    l \leq l'
    }
    {
    l \guards \textsf{int}^{l'}
    }

@@

\begin{figure*}[t]
  \begin{mathpar}
    \infer[V-Int]
    { }
    {\typeRule{\Gamma}{k}{\textsf{int}^{*}}}

@@

    \infer[V-Sub]
    {
    \typeRule{\Gamma}{v}{t'} \\
    t' \leq t
    }
    {\typeRule{\Gamma}{v}{t}}
  \end{mathpar}
  \caption{Typing judgements for values in \langName.}
  \label{fig:value_types}
\end{figure*}

@@

\begin{figure*}[t]
  \begin{mathpar}
    \infer[E-Value]
    {
    \typeRule{\Gamma}{v}{t}
    }
    {
    \typeRule{*, \Gamma}{v}{t}
    }

@@

    \infer[E-Bracket]
    {
    \typeRule{pc \sqcup pc', \Gamma}{e_1}{t} \\
    \typeRule{pc \sqcup pc', \Gamma}{e_2}{t} \\
    pc' \in H \\
    pc' \guards t
    }
    {
    \typeRule{pc, \Gamma}{<e_1 | e_2>}{t}
    }
  \end{mathpar}
  \caption{Typing judgements for expressions in \langName.}
  \label{fig:expr_types}
\end{figure*}

@@

  As a result, we know that $e$ reduces to an answer, because otherwise it would
  be stuck and this would mean that one of the projections is also stuck by
  Lemma~\ref{lem:stuck_bracket}, a contradicton.  Thus, we know that if both
  projections step to an answer, so must the original expression.
\end{proof}

@@

\begin{lemma}[Substitution]
  \label{lem:substitution}
  Assume $\typeRule{ }{v}{s}$.  Then, $\typeRule{\Gamma[x \mapsto s]}{v'}{t}$
  implies $\typeRule{\Gamma}{v'[x \Leftarrow v]}{t}$.  Also $\typeRule{pc,
  \Gamma[x \mapsto s]}{e}{t}$ implies $\typeRule{pc, \Gamma}{e[x \Leftarrow
  v]}{t}$.
\end{lemma}
\begin{proof}
  We can prove both statements by induction on the typing derivation for $v'$
  and $e$.  In this proof, we focus only on a subset of the cases.

@@

Our proof of Lemma~\ref{lem:soundness} actually turned out to be instructive
because it forced us to recognize that the original formulation in Simonet and
Pottier's paper on \origLang~\cite{InfoFlowML} wasn't quite correct.
Specifically, it is entirely possible that a step in the operational semantics
does not translate to a step in the operational semantics under either
projection, as in the case of the \ref{eq:lift-app} rule, where $\lift{e}_i$ and
$\lift{e'}_i$ are identical.  Luckily, we are fairly confident, given that we
were still able to prove Lemma~\ref{lem:stuck_bracket} despite this change, that
this is not a critical error in the original paper.
